---
title: "Steam Games Dataset"
author: "Mykyta Khrabust"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Libraries

```{r libraries}
library(readr)
library(dplyr)
library(lubridate)
library(tidyr)
```

## Helper function

Parse Python-style bracket lists **and** plain comma-separated strings into R character vectors.

```{r helper}
# "['English', 'French']"  →  c("English", "French")
# "RPG,Story Rich,Indie"   →  c("RPG", "Story Rich", "Indie")
parse_list_column <- function(x) {
  lapply(x, function(val) {
    if (is.na(val) || val == "" || val == "[]") return(character(0))

    if (startsWith(trimws(val), "[")) {
      # Python bracket-style list: strip brackets, extract quoted tokens
      inner   <- gsub("^\\[|\\]$", "", trimws(val))
      matches <- regmatches(inner, gregexpr("'([^']*)'", inner))[[1]]
      items   <- gsub("^'|'$", "", matches)
    } else {
      # Plain comma-separated string
      items <- strsplit(val, ",")[[1]]
    }

    trimws(items)
  })
}
```

## Read raw CSV

All columns are read as `character` first to avoid silent type coercion.

```{r read}
raw <- read_csv(
  "games_trimmed.csv",                                  
  col_types   = cols(.default = col_character()),
  show_col_types = FALSE
)
```

```{r}
cat(nrow(raw), "rows ×", ncol(raw), "columns\n")
```

## Parse & clean

Values in Estimated.owners are kept as-is as they are just meant to represent vague groups

```{r}
unique(raw$Estimated_owners)
```

```{r parse}
steam <- raw %>%

  # --- Platform columns: character → logical (TRUE / FALSE) ------------------
  mutate(
    Windows = as.logical(Windows),
    Mac     = as.logical(Mac),
    Linux   = as.logical(Linux)
  ) %>%

  # --- Release date: "Oct 15, 2019" → Date -----------------------------------
  mutate(
    `Release_date` = mdy(`Release_date`)
  ) %>%

  # --- Numeric columns -------------------------------------------------------
  mutate(
    across(
      c(`Peak_CCU`, Price, Discount, `DLC_count`,
        `Metacritic_score`, `User_score`,
        Positive, Negative, `Score_rank`,
        Achievements, Recommendations,
        `Average_playtime_forever`, `Average_playtime_two_weeks`,
        `Median_playtime_forever`,  `Median_playtime_two_weeks`),
      as.numeric
    )
  ) %>%

  # --- List-columns ----------------------------------------------------------
  # Each cell becomes a character vector; the column becomes a list-column
  mutate(
    `Supported_languages`  = parse_list_column(`Supported_languages`),
    `Full_audio_languages` = parse_list_column(`Full_audio_languages`),
    Developers             = parse_list_column(Developers),
    Publishers             = parse_list_column(Publishers),
    Categories             = parse_list_column(Categories),
    Genres                 = parse_list_column(Genres),
    Tags                   = parse_list_column(Tags)
  )
```

## Column types

```{r types}
data.frame(
  column = names(steam),
  type   = sapply(steam, function(x) paste(class(x), collapse = " / "))
)
```

## Working with list-columns

### Tag count per game

```{r tag-count}
steam <- steam %>%
  mutate(n_tags = lengths(Tags))

steam %>%
  select(Name, n_tags) %>%
  arrange(desc(n_tags)) %>%
  head(10)
```

### Filter by tag

```{r filter-tag}
steam %>%
  filter(sapply(Tags, function(t) "RPG" %in% t)) %>%
  select(Name, `Release_date`, Price, n_tags) %>%
  head(10)
```

### Unnest to long format (for frequency analysis)

```{r unnest, eval=FALSE}


steam_tags_long <- steam %>%
  select(AppID, Name, Tags) %>%
  unnest(Tags)

# Most common tags across the dataset
steam_tags_long %>%
  count(Tags, sort = TRUE) %>%
  head(20)
```
