---
title: "Steam Games Dataset"
author: "Mykyta Khrabust, "
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Libraries

```{r libraries}
library(readr)
library(dplyr)
library(lubridate)
library(tidyr)
library(forcats)
library(ggplot2)
```

------------------------------------------------------------------------

# Parsing the table

## Helper function

Parse Python-style bracket lists **and** plain comma-separated strings into R character vectors.

```{r helper}
# "['English', 'French']"  →  c("English", "French")
# "RPG,Story Rich,Indie"   →  c("RPG", "Story Rich", "Indie")
parse_list_column <- function(x) {
  lapply(x, function(val) {
    if (is.na(val) || val == "" || val == "[]") return(character(0))

    if (startsWith(trimws(val), "[")) {
      # Python bracket-style list: strip brackets, extract quoted tokens
      inner   <- gsub("^\\[|\\]$", "", trimws(val))
      matches <- regmatches(inner, gregexpr("'([^']*)'", inner))[[1]]
      items   <- gsub("^'|'$", "", matches)
    } else {
      # Plain comma-separated string
      items <- strsplit(val, ",")[[1]]
    }

    trimws(items)
  })
}
```

## Read raw CSV

All columns are read as `character` first to avoid silent type coercion.

```{r read}
raw <- read_csv(
  "games_trimmed.csv",                                  
  col_types   = cols(.default = col_character()),
  show_col_types = FALSE
)
```

```{r}
cat(nrow(raw), "rows ×", ncol(raw), "columns\n")
```

## Parse & clean

Values in Estimated.owners are kept as-is as they are just meant to represent vague groups

```{r}
unique(raw$Estimated_owners)
```

```{r parse}
steam <- raw %>%

  # --- Platform columns: character → logical (TRUE / FALSE) ------------------
  mutate(
    Windows = as.logical(Windows),
    Mac     = as.logical(Mac),
    Linux   = as.logical(Linux)
  ) %>%

  # --- Release date: "Oct 15, 2019" → Date -----------------------------------
  mutate(
    `Release_date` = mdy(`Release_date`)
  ) %>%

  # --- Numeric columns -------------------------------------------------------
  mutate(
    across(
      c(`Peak_CCU`, Price, Discount, `DLC_count`,
        `Metacritic_score`, `User_score`,
        Positive, Negative, `Score_rank`,
        Achievements, Recommendations,
        `Average_playtime_forever`, `Average_playtime_two_weeks`,
        `Median_playtime_forever`,  `Median_playtime_two_weeks`),
      as.numeric
    )
  ) %>%

  # --- List-columns ----------------------------------------------------------
  # Each cell becomes a character vector; the column becomes a list-column
  mutate(
    `Supported_languages`  = parse_list_column(`Supported_languages`),
    `Full_audio_languages` = parse_list_column(`Full_audio_languages`),
    Developers             = parse_list_column(Developers),
    Publishers             = parse_list_column(Publishers),
    Categories             = parse_list_column(Categories),
    Genres                 = parse_list_column(Genres),
    Tags                   = parse_list_column(Tags)
  )
```

## Column types

```{r types}
data.frame(
  column = names(steam),
  type   = sapply(steam, function(x) paste(class(x), collapse = " / "))
)
```

## Working with list-columns

### Tag count per game

```{r tag-count}
steam <- steam %>%
  mutate(n_tags = lengths(Tags))

steam %>%
  select(Name, n_tags) %>%
  arrange(desc(n_tags)) %>%
  head(10)
```

### Filter by tag

```{r filter-tag}
steam %>%
  filter(sapply(Tags, function(t) "RPG" %in% t)) %>%
  select(Name, `Release_date`, Price, n_tags) %>%
  head(10)
```

### Unnest to long format (for frequency analysis)

```{r unnest, eval=FALSE}


steam_tags_long <- steam %>%
  select(AppID, Name, Tags) %>%
  unnest(Tags)

# Most common tags across the dataset
steam_tags_long %>%
  count(Tags, sort = TRUE) %>%
  head(20)
```

## Adding % of positive reviews column

```{r}
steam <- steam %>%
  mutate(Positive_Percent = 100 * Positive / (Positive + Negative))

```

------------------------------------------------------------------------

# Part A : Basic characteristics of the dataset

## Estimated ownership

```{r}
humanize_owner_est <- function(x) {
      parts <- strsplit(x, "-")[[1]]
      format_num <- function(n) {
        n <- as.numeric(trimws(n))
        if      (n == 0)           "0"
        else if (n < 1e3)          as.character(n)
        else if (n < 1e6)          paste0(n / 1e3, "K")
        else if (n < 1e9)          paste0(n / 1e6, "M")
        else                       paste0(n / 1e9, "B")
      }
      paste0(format_num(parts[1]), "-", format_num(parts[2]))
    }
```

```{r}
# Parse the lower bound of each group to sort categories numerically
steam_owners <- steam %>%
  count(Estimated_owners, name = "n_games") %>%
  mutate(
    sort_lower_bound = as.numeric(sub("-.*", "", Estimated_owners)),
    Estimated_owners = fct_reorder(Estimated_owners, sort_lower_bound)
  ) %>%
  mutate( # converting zeros to n,m,b 20000 -> 20K
    Steam_owners_human = sapply(as.character(Estimated_owners), humanize_owner_est),
    Steam_owners_human = fct_reorder(Steam_owners_human, sort_lower_bound)
  )

```

```{r}
ggplot(steam_owners, aes(x = Steam_owners_human, y = n_games)) +
  geom_col(fill = "#7040BF", width = 0.7) +
  geom_text(
    aes(label = n_games),
    vjust = -0.5,       # place label just above the bar
    size  = 3.5,
    fontface = "bold",
    color = "#2c2c2c"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))  # leave room for labels at the top
  ) +
  labs(
    title   = "Number of games by estimated owners group",
    x       = "Estimated owners",
    y       = "Number of games"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    plot.title       = element_text(face = "bold", margin = margin(b = 12))
  )
```

```{r}
steam %>%
  filter(Estimated_owners == "100000000 - 200000000") %>%
  select(Name, Estimated_owners, Price, Release_date)
```

```{r}
steam_violin <- steam %>%
  filter(!is.nan(Positive_Percent)) %>%
  mutate(
    lower_bound = as.numeric(sub("-.*", "", Estimated_owners)),
    Estimated_owners_human = sapply(as.character(Estimated_owners), humanize_owner_est),
    Estimated_owners_human = fct_reorder(Estimated_owners_human, lower_bound)
  )

# Pre-compute mean and median per group for the overlay
ownership_stats <- steam_violin %>%
  group_by(Estimated_owners_human) %>%
  summarise(
    mean_pp   = mean(Positive_Percent),
    median_pp = median(Positive_Percent),
    .groups   = "drop"
  )

ggplot(steam_violin, aes(x = Estimated_owners_human, y = Positive_Percent)) +
  geom_violin(fill = "#45AF4B", alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.1, fill = "white", color = "#2c2c2c",
               outlier.shape = NA) +          # thin boxplot inside violin
  geom_point(data = ownership_stats,
             aes(y = mean_pp, shape = "Mean"),
             color = "#e15759", size = 2.5) +
  geom_point(data = ownership_stats,
             aes(y = median_pp, shape = "Median"),
             color = "#f28e2b", size = 2.5) +
  scale_shape_manual(values = c("Mean" = 17, "Median" = 16), name = NULL) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10)) +
  labs(
    title   = "Distribution of Positive Review % by Estimated Owners group",
    x       = "Estimated owners",
    y       = "Positive reviews (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", margin = margin(b = 10)),
    legend.position  = "top"
  )
```

## Number of games by year

```{r}
# Count games per release year, drop rows with missing dates
steam_by_year <- steam %>%
  filter(!is.na(Release_date)) %>%
  mutate(year = as.integer(format(Release_date, "%Y"))) %>%
  count(year, name = "n_games")

options(repr.plot.width = 10, repr.plot.height =2)
ggplot(steam_by_year, aes(x = year, y = n_games)) +
  geom_line(color = "#4e79a7", linewidth = 0.9) + # line
  geom_point(color = "#4e79a7", size = 2.5) + # scatter
  geom_text( # number labels 
    aes(label = n_games),
    vjust = -0.9,
    size  = 3,
    fontface = "bold",
    color = "#2c2c2c"
  ) +
  scale_x_continuous(breaks = seq(min(steam_by_year$year),
                                  max(steam_by_year$year), by = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    title = "Number of games released per year",
    x     = "Year",
    y     = "Number of games"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title     = element_text(face = "bold", margin = margin(b = 12))
  )
```

```{r}
year_min <- 2011

steam_by_halfyear <- steam %>%
  filter(!is.na(Release_date)) %>%
  mutate(
    year   = as.integer(format(Release_date, "%Y")),
    half   = ifelse(as.integer(format(Release_date, "%m")) <= 6, 1, 2),
    period = year + (half - 1) * 0.5
  ) %>%
  count(period, name = "n_games") %>%
  filter(period >= year_min, period < 2026)   # drop pre-2011 and incomplete 2026

train_half <- steam_by_halfyear %>%
  filter(period > year_min) %>%               # avoid log(0)
  mutate(
    log_games = log(n_games),
    log_t     = log(period - year_min),
    log_t2    = log_t^2
  )
# Training a model 
model_games_by_half_year <- lm(log_games ~ log_t + log_t2, data = train_half)

# Confidence intervals
pred_ci <- steam_by_halfyear %>%
  filter(period > year_min) %>%
  mutate(
    log_t  = log(period - year_min),
    log_t2 = log_t^2
  )

# predict() with interval="confidence" returns fit, lwr, upr in log space
ci_matrix <- predict(model_games_by_half_year,
                     newdata  = data.frame(log_t  = pred_ci$log_t,
                                           log_t2 = pred_ci$log_t2),
                     interval = "prediction",
                     level    = 0.95)

pred_ci <- pred_ci %>%
  mutate(
    n_games_pred = exp(ci_matrix[, "fit"]),
    ci_lwr       = exp(ci_matrix[, "lwr"]),   # back-transform to original scale
    ci_upr       = exp(ci_matrix[, "upr"])
  )

# Predictions back-transformed to original scale
pred_df <- steam_by_halfyear %>%
  filter(period > year_min) %>%
  mutate(
    log_t        = log(period - year_min),
    log_t2       = log_t^2,
    n_games_pred = exp(predict(model_games_by_half_year,
                               newdata = data.frame(log_t  = log_t,
                                                    log_t2 = log_t2)))
  )

ggplot() +
  geom_ribbon(
    data  = pred_ci,
    aes(x = period, ymin = ci_lwr, ymax = ci_upr),
    fill  = "#f28e2b", alpha = 0.2
  ) +
  geom_point(
    data = steam_by_halfyear %>% filter(period > year_min),
    aes(x = period, y = n_games),
    color = "#4e79a7", size = 2.2
  ) +
  geom_line(
    data = pred_df,
    aes(x = period, y = n_games_pred),
    color = "#f28e2b", linewidth = 1.0, linetype = "dashed"
  ) +
  scale_x_continuous(
    breaks = seq(year_min, 2025, by = 1),
    minor_breaks = seq(year_min, 2025, by = 0.5)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  labs(
    title    = "Games released per half-year — quadratic power trend (2011+)",
    subtitle = sprintf(
      "n_games = %.2f · (period − %d)^(%.2f + %.2f·log(period − %d))",
      a, year_min, p, q, year_min),
    x = "Period", y = "Number of games"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_line(color = "grey92"),
    plot.title       = element_text(face = "bold", margin = margin(b = 4)),
    plot.subtitle    = element_text(color = "#f28e2b", margin = margin(b = 10))
  )
```

```{r}
a <- exp(coef(model_half)[["(Intercept)"]])
p <- coef(model_half)[["log_t"]]
q <- coef(model_half)[["log_t2"]]

cat("Fitted model formula:\n")
cat(sprintf(
  "n_games = %.4f * (period - %d) ^ (%.4f + %.4f * log(period - %d))\n",
  a, year_min, p, q, year_min
))
```

### Diagnostyka modelu

```{r}
summary(model_games_by_half_year)

```

```{r}
confint(model_games_by_half_year)
```

```{r}
plot(model_games_by_half_year)
# par(mfrow = c(2, 2)); plot(model_games_by_half_year); par(mfrow = c(1, 1))
```

```{r}
# See which periods the outlier rows correspond to
influential <- c(1, 2, 3)   # replace with the numbers you see on the plots
train_half[influential, ]
```

### Model diagnostics summary

The model was fit in log-log space using a quadratic polynomial:

$$\log(n\_games) = a + p \cdot \log(t) + q \cdot \log^2(t), \quad t = \text{period} - 2011$$

estimated on half-yearly Steam release counts from 2011.5 to 2025.5 (n = 29).

**LINE assumptions:**

-   **Linearity.** *Residuals vs Fitted* shows the red line close to zero across most of the fitted range with no pronounced curve, suggesting the quadratic log-log transformation adequately captures the trend. A mild downward slope at the right tail indicates slight overestimation in the most recent periods.

-   **Independence.** Observations are aggregated half-year counts from a single growing platform — consecutive periods are likely positively correlated (more developers join every year). Formal independence cannot be claimed; a time-series model would be more appropriate if serial correlation is a concern.

-   **Normality.** The *Q-Q plot* shows most points tracking the diagonal well in the central range. Points 1 and 4 deviate noticeably in both tails, indicating heavier-than-normal tails. With n = 29 this is a moderate concern but not critical for inference given the large F-statistic.

-   **Equal variance (homoscedasticity).** *Scale-Location* shows a declining trend — residual spread is larger for small fitted values (early periods) and shrinks toward recent years. This is a mild violation of homoscedasticity. Point 4 (visible in all four panels) is the primary driver.

**Point 4** (identifiable via `train_half[4, ]`) appears as an influential outlier across all diagnostic plots and sits beyond Cook's distance boundary in *Residuals vs Leverage*. It corresponds to an early period with anomalously low releases relative to the fitted trend and meaningfully affects the coefficient estimates.

**Coefficients and model fit:**

All three coefficients are highly significant: the intercept (p \< 2e-16), the linear log term `log_t` (p = 5.35e-06), and the quadratic term `log_t2` (p = 5e-04). The F-statistic of 328.5 on 2 and 26 degrees of freedom confirms that the model as a whole explains the data far better than an intercept-only baseline.

**R² = 0.962** (adjusted 0.959) means the model accounts for \~96% of variance in log-transformed game counts — a strong fit for real-world count data. The residual standard error of 0.28 on the log scale translates to a typical multiplicative error of roughly e0.28≈1.32e^{0.28} \approx 1.32
e0.28≈1.32, i.e. predictions are off by about ±32% on the original scale.

**Overall conclusion:** the model fits well and all predictors are significant, but two assumptions — independence and homoscedasticity — are not fully met. Results should be interpreted as a descriptive trend model rather than a formal inferential one.
